---
title: "Using Sensor Data To Model Physical Exercise"
author: "Kamran Haroon"
date: "February 11, 2016"
output: html_document
---


### Summary
This process uses motion sensor data and machine learning tools to predict whether pariticipants perform a physical exercise correctly, or not. the training and test data was provided by . Columns not related to sensor measurements were removed from the training data set. For cross-validation, the training data was further split into training and preTest data sets. The model was built using random forest and yielded an accuracy of xx%. For final verification, our model passed 20 prediction tests on previously unseen data.


### Data Processing



#### Loading the data


```{r loadtraindata, echo = TRUE}
  set.seed(327)

  # Verify the pml-training.csv has been downloaded 
  if (!file.exists("./data/pml-training.csv")) {
    dataUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
    download.file(dataUrl, destfile = "./data/pml-training.csv", method = "curl")
  }

  # load the data in the pmltrain dataframe
  pmlTrain <- read.csv("./data/pml-training.csv", header = TRUE)
  dim(pmlTrain)
```


#### Feature Selection
```{r libload, echo = FALSE, message = FALSE}
  library(dplyr)
  library(caret)
```

```{r tidytrainingdata, echo = TRUE}
  # Verify the pml-testing.csv has been downloaded 
  if (!file.exists("./data/pml-testing.csv")) {
    dataUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
    download.file(dataUrl, destfile = "./data/pml-testing.csv", method = "curl")
  }

  # load the data in the test dataframe
  pmlTest <- read.csv("./data/pml-testing.csv", header = TRUE)
  dim(pmlTest)
  
  # get a list of the test columns with NA in the column
  naNames <- names(pmlTest[, colSums(is.na(pmlTest)) == nrow(pmlTest)])
  
  # remove columns from the training set that are NA in the test set 
  pmlTrain <- pmlTrain[,-which(names(pmlTrain) %in% naNames)]
```

```{r reducepredicters, echo = TRUE}
  # only keep the sensor based columns and the classe 
  pmlTrain <- select(pmlTrain, roll_belt:classe)
  names(pmlTrain)
```


#### Data Exploration
```{r exploredata, echo = TRUE}
  library(ggplot2)
  qplot(roll_belt, pitch_belt, colour=classe, data=pmlTrain)
```


#### Cross-validation
```{r splitdata, echo = TRUE}
  inTrain <- createDataPartition(y=pmlTrain$classe, p=0.60, list=FALSE)
  training <- pmlTrain[inTrain,] 
  preTest <- pmlTrain[-inTrain,] 
```

### Model Creaton and verificaton
``` {r trainmodel2, echo = TRUE, cache = TRUE, message = FALSE}
  # step 1: configure parallel processing
  library(parallel)
  library(doParallel)
  cluster <- makeCluster(detectCores() -1)
  registerDoParallel(cluster)
  
  # step 2: configure trainControl object
  fitControl <- trainControl(method = "cv", number = 10, allowParallel = T)
  
  # step 3: develop training model
  modFit <- train(classe~., method = "rf", data=training, trControl = fitControl)
  
  # step 4: close the cluster
  stopCluster(cluster)
```


#### Accuracy and Out-of-sample Error
```{r verifymodel, echo=TRUE, message=FALSE}
  predictions <- predict(modFit, preTest)
  confusionMatrix(predictions,preTest$classe)
```

### Project Prediction Quiz
```{r, finalpredictions, echo=TRUE, message=FALSE}
  predictfinal <- predict(modFit, pmlTest)
  predictfinal

  results <- "./results"
  if(!file.exists(results)) {dir.create(results)}

  pml_write_files = function(x,y){
    n = length(x)
    for(i in 1:n){
      filename = paste(y,"/problem_id_",i,".txt",sep="")
      write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
    }
  }

  pml_write_files(predictfinal,results)  
```